---
title: "BigQuery"
subtitle: "Centralized database for integrated expedition data"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(here)
library(knitr)
library(gt)
```

## Overview

The Pristine Seas Science Database is a centralized, modular system hosted in Google BigQuery that serves as the foundation of our data management infrastructure. This system integrates ecological data collected across more than a decade of scientific expeditions, supporting high-integrity, reproducible research on marine biodiversity and informing global ocean conservation policy.

#### Why BigQuery?

Google BigQuery offers several advantages for marine ecological data:

  - Scalability: Handles our growing dataset spanning 40+ expeditions without performance degradation
  - Integration: Seamlessly connects with our R-based analysis workflows
  - Collaboration: Enables standardized access across our distributed research team
  - Future-proofing: Provides a robust platform that can evolve with our research needs

Note: BigQuery is were Global Fishing Watch data lives and we have access to the raw backend data. 

#### FAIR Data Principles

Our database implementation adheres to the FAIR data principles:

::: {.grid}
::: {.g-col-3}
#### Findable
- Unique and unified IDs
- WoRMS-linked taxonomy
- Consistent spatial hierarchy
:::

::: {.g-col-3}
#### Accessible
- Query-ready BigQuery tables
- Controlled access and permissions
- Varied connection options
:::

::: {.g-col-3}
#### Interoperable
- Spatial reference system
- Harmonized taxonomic backbone
- Consistent units (cm, g, m²)
:::

::: {.g-col-3}
#### Reusable
- Comprehensive metadata
- Reproducible code
- Clear provenance6
:::
:::

## Database Architecture

The Pristine Seas Science Database is organized around two major dataset groups that balance flexibility with consistency:

1. **Method Datasets**: Specific to each survey technique (UVS, BRUVS, eDNA, etc.)
2. **Reference Datasets**: Shared taxonomic, spatial, and lookup tables providing a unified backbone

### Database Organization

```{mermaid}
graph LR
    %% Main project node
    pristine["pristine-seas"] 
    
    %% First level - dataset categories
    reference_group["Reference Datasets"]
    method_group["Method Datasets"]
    
    %% Reference datasets
    exp["expeditions/"]
    tax["taxonomy/"]
    look["lookup/"]
    
    %% Method datasets
    uvs["uvs/"]
    pbruv["pbruv/"]
    sbruv["sbruv/"]
    edna["edna/"]
    other["Other Methods..."]
       %% UVS tables
    sites["sites"]
    stations["stations"]
    
    %% Fish transect tables
    blt_group["Fish Belt Transect Tables"]
    blt1["blt_stations"]
    blt2["blt_observations"]
    blt3["blt_biomass_by_taxa"]
    
    %% LPI tables
    lpi_group["Benthic LPI Tables"]
    lpi1["lpi_stations"]
    lpi2["lpi_counts"]
    lpi3["lpi_cover_by_taxa"]
    
    %% Taxonomy tables
    fish["fish"]
    benthos["benthos"]
    inverts["inverts"]
    
    %% Expeditions tables
    info["info"]
    exp_sites["sites"]
    
    %% Connections
    pristine --> reference_group
    pristine --> method_group
    
    reference_group --> exp
    reference_group --> tax
    reference_group --> look
    
    method_group --> uvs
    method_group --> pbruv
    method_group --> sbruv
    method_group --> edna
    method_group --> other
    
    uvs --> sites
    uvs --> stations
    uvs --> blt_group
    uvs --> lpi_group
    
    blt_group --> blt1
    blt_group --> blt2
    blt_group --> blt3
    
    lpi_group --> lpi1
    lpi_group --> lpi2
    lpi_group --> lpi3
    
    tax --> fish
    tax --> benthos
    tax --> inverts
    
    exp --> info
    exp --> exp_sites
    
    %% Styling with improved visibility and contrast
    classDef root fill:#004165,color:#ffffff,stroke:#002e48,stroke-width:2px,rx:8,ry:8
    classDef group fill:#f8f9fa,color:#000000,stroke:#343a40,stroke-width:1px,stroke-dasharray: 5 5,rx:5,ry:5
    classDef refDataset fill:#d4edda,color:#000000,stroke:#28a745,stroke-width:2px,rx:8,ry:8
    classDef methodDataset fill:#cce5ff,color:#000000,stroke:#0d6efd,stroke-width:2px,rx:8,ry:8
    classDef table fill:#ffffff,color:#000000,stroke:#6c757d,stroke-width:1px,rx:3,ry:3
    classDef tableGroup fill:#f8f9fa,color:#000000,stroke:#343a40,stroke-width:1px,rx:3,ry:3
    
    class pristine root
    class reference_group,method_group group
    class exp,tax,look refDataset
    class uvs,pbruv,sbruv,edna,other methodDataset
    class sites,stations,info,exp_sites,blt1,blt2,blt3,lpi1,lpi2,lpi3,fish,benthos,inverts table
    class blt_group,lpi_group tableGroup
```

For the complete database documenation please refer to the [Pristine Seas Database Documentation](https://pristine-seas.github.io/legacy-db/).

### Common Dataset Structures

Each method dataset follows a consistent pattern with tables organized into logical categories that support various analytical needs. This standardized structure enables efficient querying, cross-method integration, and reproducible science.

::: {.callout-note appearance="simple"}

## Standard Table Types

- **Site Tables** (`sites`, `[method]_sites`)
  - One row per survey site or deployment
  - Contains spatial information, geographic coordinates, habitat descriptions, and context
  - Key tables: 
    - `expeditions.sites`: aggregates all survey sites across methods and expeditions over time
    - `uvs.sites`: all underwater visual survey sites with method specific metadata.
    - `sub.sites`: all sub dive sites with method specific metadata.

- **Station Tables** (`[method]_stations`)
  - One row per sampling unit (e.g., depth strata, replicate)
  - Includes spatial and temporal information, survey effort, and method-specific metadata
  - Include aggregated metrics of interest (e.g., fish biomass, % coral cover)
  - Functions as a primary unit for analysis and comparison
  - Key tables: 
    - `uvs.lpi_stations`: all benthic LPI stations, including transect length, depth, habitat type, and summary metrics.
    - `uvs.blt_stations`: all fish belt transect stations, including transect length, depth, habitat type, and summary metrics. 
    - `sub.stations`: all submersible survey stations (horizontal transects) done with the sub.
  
- **Observation Tables** (`[method]_observations`)
  - Contains individual records (e.g., fish counts, benthic points, video annotations)
  - Stores raw QAQC'd ecological data with validated taxonomic identifications
  - Key tables: 
    - `uvs.lpi_counts`: all benthic LPI counts, including taxonomic identifications (`field_name`, `morphotaxa`, `accepted_aphia_ID`).
    - `uvs.blt_observations`: all fish belt transect observations, including taxonomic identifications and estimated biomass.

- **Station-Taxa Tables** (`[method]_[metric]_by_[dimension]`)
  - Aggregated analysis-ready metrics
  - Pre-calculated to standardize common analytical outputs
  - Enables efficient cross-site and cross-method comparisons
  - Key tables: 
    - `uvs.cover_by_taxa`: total point and % cover by morphotaxa and station.
    - `uvs.biomass_by_taxa`: fish abundance and biomass by taxa and station.
    - `pbruvs.Maxn_by_station`: Max N and length estimates per taxa per station
:::


## Data Flow: Field to Database Pipeline

The integration of expedition data into BigQuery follows a standardized workflow that ensures data quality and consistency:

```{mermaid}
flowchart LR
    subgraph Field[Field]
        direction TB
        A[Collection] --> B[Validation]
        B --> C[Ship Storage]
    end
    
    subgraph Process[Processing]
        direction TB
        D[Drive Backup] --> E[Pipeline Processing]
        E --> F[Quality Control]
    end
    
    subgraph Database[Database]
        direction TB
        G[BigQuery Ingestion] --> H[Analysis-Ready Data]
    end
    
    Field --> Process
    Process --> Database
    
    classDef field fill:#004165,color:#ffffff,stroke-width:1px
    classDef process fill:#8EBDC8,color:#000000,stroke-width:1px
    classDef db fill:#E63946,color:#ffffff,stroke-width:1px
    
    class Field,A,B,C field
    class Process,D,E,F process
    class Database,G,H db
```

#### Process Steps

1. **Field Collection**: Researchers collect and record data using standardized methods and digital fieldbooks
2. **Initial Validation**: Field-level data checks ensure completeness and quality
3. **NAS Storage**: All expedition data is securely stored on the ship's Network Attached Storage
4. **Google Drive Backup**: Post-expedition, data is organized and backed up to Google Drive
5. **Pipeline Processing**: Each method's data undergoes standardized processing via code in expedition repositories
6. **Quality Assurance**: Automated and manual checks verify data integrity
7. **Database Ingestion**: Processed data is ingested into the appropriate BigQuery tables
8. **Analysis**: Data becomes available for standardized analysis workflows

## Access and Use

The Pristine Seas Science Database is designed to be accessible through multiple interfaces.

### Rstudio

The most common way to interact with our BigQuery database is through R, using the familiar tidyverse workflow.

#### Establishing a Connection

Setting up a connection is straightforward:

```{r include = T, eval = T, echo = T, code-fold: false}
#| code-fold: false
# Load required packages
library(DBI)
library(bigrquery)
library(tidyverse)

# Create a database connection
bq_connection <- DBI::dbConnect(bigrquery::bigquery(),
                      project = "pristine-seas",
                      billing = "pristine-seas")
```

This code establishes a connection to the entire database, allowing you to explore datasets and tables directly from the **Connections** pane in RStudio. The first time you run this code, you'll be prompted to authenticate with your Google account.
 
#### Using dplyr Verbs

The real power comes from using familiar dplyr verbs directly with BigQuery tables—no SQL knowledge required:

```{r include = T, eval = T, echo = T}
#| code-fold: false

df <- tbl(bq_connection, "taxonomy.fish") |> 
  group_by(family, genus) |> 
  summarize(n_taxa = n_distinct(accepted_aphia_id),
            .groups = "drop") |> 
  arrange(desc(n_taxa)) |> 
  head(30) |> 
  collect()
  
df |> 
  ggplot()+
  geom_col(aes(x = fct_reorder(family, n_taxa, sum), 
               y = n_taxa, 
               fill = genus), 
           show.legend = T)+
  theme_minimal()+
  coord_flip()+
  labs(x = "", y = "# Distinct Taxa", fill = "",
       title = "Number of fish taxa in taxonomy.fish table",
       subtitle = "By family and genus")+
  paletteer::scale_fill_paletteer_d("ggsci::default_igv")
```

What makes this approach powerful:

1. **BigQuery does the heavy lifting** – All filtering, grouping, and summarizing happens in the database
2. **Only results are transferred** – Data never loads into memory until you call `collect()`
3. **Familiar syntax** – The same dplyr verbs you already use with local data frames
4. **Readable code** – Complex queries expressed in clean, maintainable R code


### Google BigQuery Console

The database can also be accessed directly through the Google BigQuery Console:

1. Visit [console.cloud.google.com/bigquery](https://console.cloud.google.com/bigquery)
2. Navigate to the `pristine-seas` project
3. Use the query editor to write and execute SQL queries
4. Explore tables, schemas, and query history

The console provides a user-friendly interface for exploring table structures, examining data samples, and running ad-hoc queries without writing code.

### Access Management

Access to the Pristine Seas Science Database is managed through Google Cloud IAM:

- **Team Members**: Full read access to all datasets
- **Collaborators**: Read access to specific datasets relevant to their work
- **Partners**: Access via shared exports or temporary read credentials
- **Public**: Access to published, non-sensitive data via data packages (still TBD)

To request access, contact the database administrator with your Google account email and purpose.

### Best Practices

When working with the Pristine Seas Science Database:

1. **Minimize data transfer**: Filter data in BigQuery before collecting to R
2. **Use primary keys**: Join tables using established keys (`ps_station_id`, `ps_site_id`)
3. **Reproducible queries**: Document your queries in Quarto documents
4. **Analysis patterns**: Build on established workflows in expedition repositories

## Maintenance

The Pristine Seas Science Database is actively being developed and continously improved to meet the needs of our team. 
